Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-10-31 Paper/Paper.tex

\begin{abstract}
    Technological innovation has affected workers and labor markets for decades. I add to the understanding of technology implementation's affect on highly technical jobs by investigating changes in physician behavior as a result of a major technology shift in U.S. health care, electronic health records (EHRs). I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital and estimate average group time treatment effects on various labor market outcomes. I find that physicians are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHR exposure. Further, for physicians who remain exposed to the technology, the number of patients seen increases. These results point to unintended consequences of rapid technology implementation in technical occupations, but also indicates potential productivity gains for those who embrace the technology.
\end{abstract}

@@

Physicians, as the primary users of EHRs, play an important role in whether the expected benefits come to fruition. The daily tasks a physician faces change drastically when a new system is put in place. Some physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHR.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards alternative work settings. However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question. Further, the cost imposed is likely to vary based on the type of worker, in particular, age group. Using a difference-in-differences research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices.

I form a panel of physicians who work in hospitals, hereafter referred to as hospitalists or physicians interchangeably, which spans from 2009-2017. Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify hospital-physician pairs that indicate the physician is working within the hospital. Then, I use the American Hospital Association (AHA) Hospital Survey to determine which hospitals implement an EHR, and thus connect hospitalists to EHR use. Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for detailed information on the patients seen by each hospitalist. For all outcomes, the independent variable of interest is a binary treatment variable capturing exposure to an EHR. I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing patients in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting and location of the physician office, (3) number of patients seen, and (4) claims filed per patient.

This paper contributes to the literature's understanding of how health information technology generally affects health care. There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use. Despite a large number of case studies that find generally positive effects in improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), empirical work has found a mixture of results: that outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}). More relevant to individual physicians' productivity, a number of studies consider the effect of EHRs on productivity in specific settings. One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EHRs being adopted in primary-care sites (\cite{Meyerhoefer}).

I expand this literature, first by considering a different mechanism by which EHRs could affect health care through physicians, which, to my knowledge, has not yet been examined. This is the first empirical study that connects EHR implementation to physician retirement and practice location, and focuses on the difference in response across age groups. Further, I improve on the health information technology literature by considering the primary time period in which EHRs were rapidly implemented. This provides sufficient variation in the timing of implementation and ensures the EHRs are advanced enough to be considered ``meaningful" and are likely affecting physicians' daily life. Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework. I improve on this strategy by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation with heterogeneous effects, and further provides insight to short vs. long term effects.

Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}). Contrary to this result, I find that EHR implementation led to a 0.006 ppt increase in the likelihood of a senior physician retiring after implementation, a 20\% increase relative to the average retirement for this age group. Physicians less than 60 were 0.002 ppts more likely to retire after exposure, possibly pointing to career switching. These results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence early career physicians. For those who did not retire, I find that young physicians were .06 ppts (15\% relative to the mean) more likely to see patients in an office setting the same year as EHR exposure. I find that patient count increases by at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who retired.

Two key assumptions underlie the main analysis. First, I assume that physicians are constrained to use the technology implemented in their work setting. To test this assumption, I investigate the existence of hospital employees with the purpose of using an EHR on behalf of a physician, which I refer to as data assistants. If data assistants are used to ease EHR use for physicians, the results are at least partially driven by their existence. I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the vast amount of EHR implementation. Further, my main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results. Second, I assume the decision for hospitals to implement EHRs is exogenous to individual physicians. That is, a physician does not choose their own EHR exposure. While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making. For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a low likelihood that EHR implementation is endogenous. The results of the additional analyses indicate that my main findings are not sensitive to endogeneity concerns stemming from joint EHR decision making among physicians and hospitals.

@@

\subsection{Background on EHRs}
EHRs have been an important feature of health care since the 1980s. Early in the technology's existence, health care professionals perceived the technology as a complement to paper records, primarily deployed by large academic medical centers to improve efficiency in billing and/or scheduling. Physicians did not interact directly with these early-generation EHRs, and thus were not drastically affected by their implementation. As technological innovation made computers more portable, the usability of EHRs increased, creating what is known as the ``physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates. Despite usability, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry. Automation in data entry was non-existent, making it extremely time consuming for the user. Even when automation was developed for particular machines that performed monitoring, the hospital still held responsibility for the accuracy of information and therefore required physicians to manually check each data entry (\cite{evans2016electronic}).

The HITECH Act was passed in 2009, designating \$27 billion in government subsidies to entities who used EHRs according to certain guidelines. The guidelines included having at least 80\% of patients in the system, regularly recording answers to specific questions, and protecting the security of the system to ensure privacy. The U.S. allocated subsidies according to stages of "meaningful" use: stage 1 (2011) focused on data collection, stage 2 (2012) extended to using the EHR for care support, and stage 3 (2014) extended to data sharing between practices. This program was successful in  spurring EHR adoption, shown in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}). Figure \ref{fig:meanuse} shows a geographical comparison of U.S. hospitals who have received stage 1 of the meaningful use subsidy in 2011 vs. 2013, revealing the nationwide and rapid expansion of technology.

@@

Figure \ref{fig:EPIC} shows two computer screenshots of what physicians might see in their EHR system. On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively (\cite{arndt2017tethered}). These are the most time consuming portions of EHR use, according to the same study. In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR. Despite a large amount of time spent with them, physicians still continue to report frustration over EHR use. The most time consuming functions are also reported as the most frustrating (\cite{dymek2021building}). A physician still likely spends almost 2 hours a day managing their inbox: deleting duplicate messages, sifting through messages meant for other members of a care team, and searching relevant information (\cite{dymek2021building}). Another common issue physicians face is the lack of usability of EHR systems, that functions either take minutes to locate or even longer to load. Hospitals often provide training upon adoption, but rapid system upgrades and changes make familiarity with the system difficult. Physicians have been voicing concerns over EHR usability since their beginning, indicating that they are likely responding in their behavior as well.

@@

This section serves to outline the potential responses of physicians who work primarily in hospitals to EHRs based on the costs and incentives they face. The physician behaviors that I consider in this analysis are closely related and should not be taken as independent. For example, a physician's choice of work setting depends on their previous decision to prolong retirement, and billing activity depends on work setting. Thus, I discuss the behaviors in stages of decision making.

@@

Workers in developed countries tend to plan for formal retirement well in advance. Generally, an exogenous shock to a worker's environment is not expected to change retirement age due to the amount of wealth and planning necessary to formally leave the labor force. Classic retirement models structure the decision of retirement by maximizing expected lifetime utility based on future earnings and retirement benefits, or choosing to prolong retirement only if the expected earnings gain from doing so is positive (\cite{gustman1986disaggregated}, \cite{stock1990pension}). However, physicians typically make this decision differently than workers in other industries. Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}). When the time to retire comes, many physicians hesitate to abandon patients they have seen over the course of their career. The decision to delay retirement is not financial, but altruistic. Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.

It is important to note that age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently. First, I consider physicians already of retirement age when a new technology is implemented. While payment schemes, wealth, and retirement benefits certainly impact the utility a physician receives from working or retiring, I assume the decision to retire is independent of those things since, on average, those in this age group have already prepared financially to retire. Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility from working depends on job satisfaction and personal life factors. Exogenous EHR implementation affects the decision to retire through its affect on job satisfaction. If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work. If EHRs make daily tasks easier, there is a positive relationship. Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction. Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).

Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role. There are opportunities for physicians of any age to switch careers towards administrative roles, teaching, consulting, or hospital management. While it is unlikely that early career physicians are formally retiring, physicians of any age could be career switching. This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how an EHR impacts job satisfaction of current clinical work. As I discuss in Section \ref{sec:data}, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch. Thus, the term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards. A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.

@@

For most physicians, EHRs will not impose a large enough cost (or benefit) to change the duration of being in the labor market. However, there are other ways physicians might change their behavior under drastic changes to their current working environment. For those who remain practicing in a clinical environment, changing the physical place of work may allow the physician more control over which technologies they use. Generally, physicians may have the ability to change hospitals, move from a hospital to an office, or work in multiple facilities and change the amount of work done in different facilities. I only consider physicians who are classified as a primary care physician and are closely tied to at least one hospital, often referred to as hospitalists or internists. \footnote{For a detailed discussion of included physicians, see Appendix \ref{app:data}} Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both. While I do not have sufficient data to observe the type of hospitalist each physician is, I investigate the incentives faced by each type.

First, I consider those who are in a relatively binding contract with the hospital. For hospitalists under full time contract with a medical practice, the physician is more tethered to the hospital. This could impact the response to EHR implementation in that hospital in two ways: (1) through potentially increased use of the EHR in its initial phases, and (2) limited access to other work settings. A typical contract requires 90 to 120 days termination notice and some contracts include do not compete clauses (\cite{yasgur_by_-_yasgur_2016}), making a shift in work setting costly or excluding it from the choice set. Since these hospitalists are established in their hospital, they work a significant amount of time there. When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings. Thus, I classify these physicians as "sticky", since they are not able to change behavior in the short run.

@@

There are likely still physicians who are not induced to change their behavior on the basis of retirement or work setting due to EHR implementation. A lack of response could be due to 4 factors: the hospitalist is bound to their current work environment, EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment. For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect output in current work environment. An important aspect of the implementation of EHRs is whether or not they allow users to be more productive. A main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.

Whether EHRs affect productivity in terms of patients seen is theoretically ambiguous. If they serve their purpose to decrease administrative burden, the physician would be freed up to see more patients. In this case, productivity would increase. However, if utilizing an EHR actually imposes an additional administrative burden as suggested in physician interviews, the number of patients could either decrease or stay the same, indicating a decrease in productivity.

Further, there is a unique relationship between EHR use and efficiency in terms of billing activity done by the physician. Billing activity refers to the number of claims filed per patient seen. EHRs may improve efficiency if they remove the need for repeated or excess testing. However, if having an EHR makes it easy for a physician to claim for additional items with the click of a button, the adverse incentive may lead physicians to claim for more items per patient. Thus, both productivity and efficiency are empirical questions.

@@

Using various linked data sets, I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, outcomes related to physician practice location and billable activity, and other relevant physician, hospital, and practice characteristics. I describe the different data sets used to construct the panel below. I include a detailed outline of data and variable construction in Appendix \ref{app:data}.

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Included in the MD-PPAS is physician NPI, primary and secondary specialties, both patient and total billing counts, and fraction of patients in different settings such as office, inpatient or emergency room. I first limit the sample of physicians based on specialty type, since my analysis relies on EHR implementation in hospitals. I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, general practice or family practice in at least one year. However, I exclude physicians who list themselves as a specialist in all but one year. Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 70\% of patients in a hospital setting in all years, and I drop physicians with insufficient claim counts to show indications of behavior.

This data includes information on physicians which I use to construct the dependent variables for my analysis. Generally, these include the decision to retire (leave clinical work), the decision to switch to a different work setting, number of patients seen, and overall billing activity. I describe these outcome variables and their construction in detail in Section \ref{sec:outcome}.

CMS Shared Patient data records annual information on count of Medicare patients and frequency of billing associated with the same patient across providers within a given time period. These data are available from 2009 to 2015. For example, if a primary care physician refers a patient to a specialist, then those two physicians have that particular shared patient in common. The number of shared patients are collected in 30, 90, or 180 day intervals. I focus on the 30 day interval, and I employ this data to detect physicians who work closely with hospitals. I limit the entities by tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS. I am specifically interested in primary care physicians who have a close working relationship with hospitals, who do rounds within at least one hospital. The types of physicians are already limited to primary care, but there still may be office-based primary care physicians in the shared patient data. Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data. I only include physician-hospital pairs who have non-missing data with at least 30 shared patients per year, which gives a threshold of work relationship without excluding physicians who have no claims for a number of years.

Using hospital National Provider Identifier (NPI), I link the physician-hospital pairs from the shared patient data to the American Hospital Association (AHA) survey, which contains information on hospital-level EHR use and other characteristics. I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with. A hospital can answer the question of whether they use an EHR with either ``Partially", ``Fully", ``Not at all". Since partial implementation of an EHR does not guarantee sufficient use, I define a hospital as implementing an EHR the first year they answer ``Fully". I then aggregate this data to the physician level. Since the Shared Patient data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015 as I do not observe whether they become treated in 2016 or 2017. That is, the data contains no units which are "never-treated".

Finally, I include a physician's graduation year from Physician Compare. I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.

@@

The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I loosely call retirement. This variable is constructed as follows: 
$$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$. Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future. \footnote{Alternatively, I could also define retirement using future number of patients. I use claim count for a more conservative measure of retirement, but using patient count yields identical results. In Appendix \ref{sec:changes}, I limit the sample years to 2009-2015 to assess the robustness of the main retirement definition.}

>>>>

\begin{abstract}
    Technological innovation has affected workers and labor markets for decades. I add to our understanding of technology's affect on highly technical jobs by investigating changes in physician behavior as a result of a major technology shift in U.S. health care, electronic health records (EHRs). I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital and estimate average group time treatment effects on various labor market outcomes. I find that physicians are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHR exposure. Further, physicians exposed to EHRs see more patients than those who are not. These results point to unintended consequences of rapid technology implementation in technical occupations, but also indicate potential productivity gains for those who embrace the technology.
\end{abstract}

@@

Physicians, as the primary users of EHRs, play an important role in whether the expected benefits come to fruition. Daily tasks change drastically when a new system is put in place; some physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHR.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards alternative work settings. However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question. Using a difference-in-differences research design, I estimate the effect of hospital EHR implementation on individual physician labor market choices, and I highlight the differences in effects for older vs. younger physicians.

I form a panel of physicians who work in hospitals, hereafter referred to as hospitalists or physicians interchangeably, which spans from 2009-2017. Using Centers for Medicare and Medicaid Services (CMS) Shared Patient Data, I identify hospital-physician pairs. Then, I use the American Hospital Association (AHA) Survey to determine which hospitals implement an EHR, and thus connect hospitalists to EHR use. Finally, I use Medicare Data on Provider Practice and Specialty (MD-PPAS) for detailed information on the patients seen by each hospitalist. The independent variable of interest is a binary treatment variable capturing exposure to an EHR. I estimate group time treatment effects of EHR exposure on the following physician decisions: (1) retirement, measured based on zero or missing patients in all future years of the panel, (2) where to physically work, measured by fraction of patients seen in an office setting and location of the physician office, (3) number of patients seen, and (4) claims filed per patient.

This paper contributes to the literature's understanding of how health information technology affects health care. There is robust empirical research examining the effect of EHRs on patient outcomes and hospital costs, but these studies largely use data prior to the subsidization of meaningful EHR use. Despite a large number of case studies that find generally positive effects in improved patient outcomes and decreased cost (surveyed in \cite{Buntin2011TheResults}), empirical work has found a mixture of results: outcomes for median patients do not change as a result of EHRs (\cite{Agha2014TheCare}, \cite{McCullough2016HealthCoordination}, \cite{Meyerhoefer}) while newborns and severe patients experience improvement in health outcomes (\cite{Miller2009}, \cite{Freedman2015}, \cite{McCullough2016HealthCoordination}), and that hospital costs only decrease 6 years after implementation, if at all (\cite{Agha2014TheCare}, \cite{dranove2014trillion}). More relevant to individual physicians' productivity, a number of case studies consider the effect of EHRs on productivity. One study finds that nursing home productivity increases after adopting health IT (\cite{Hitt2016}), but another finds that physician productivity decreases by 11 percent due to EHRs being adopted in primary-care sites (\cite{Meyerhoefer}).

I expand this literature, first by considering physicians as the mechanism by which EHRs do not affect health care, which, to my knowledge, has not yet been examined. This is the first empirical study that connects EHR implementation to physician retirement and practice location, and focuses on the difference in response across age groups. Further, I improve on the health information technology literature by considering the primary time period in which EHRs were rapidly implemented. This provides sufficient variation in the timing of implementation and ensures the EHRs are meaningful and are likely affecting physicians' daily life. Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-difference framework. I improve on this strategy by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation with heterogeneous effects, and further provides insight to short vs. long term effects.

Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}). Contrary to this result, I find that EHR implementation led to a 20\% increase relative to the average retirement for this age group. Physicians less than 60 were 7\% more likely to retire after exposure, possibly pointing to career switching. For those who did not retire, I find that physicians were 30\% more likely to see patients in an office setting the same year as EHR exposure. I find that patient count increases by at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who retired.

Two key assumptions underlie the main analysis. First, I assume that physicians are constrained to use the technology implemented in their work setting. To test this assumption, I investigate the existence of hospital employees with the purpose of using an EHR on behalf of a physician, which I refer to as data assistants. If data assistants are used to ease the burden of EHR use for physicians, the results are at least partially driven by their existence. I find that the majority of hiring of data assistants took place in 2013-2014, a two year delay from the vast amount of EHR implementation. My main results are not sensitive to the exclusion of these years, and estimating the effect of EHR implementation only in hospitals without data assistants yields noisy, but similar, results. Second, I assume the decision for hospitals to implement EHRs is exogenous to individual physicians. That is, a physician does not choose their own EHR exposure. While this assumption is reasonable in many settings, there may be instances where a physician is involved in hospital decision making. For this reason, I also consider a sensitivity analysis in which I limit the sample only to hospitals where vertical integration between physician and the hospital is low, as this indicates a low likelihood that EHR implementation is endogenous. The results of the additional analyses indicate that for most outcomes, my main findings are not sensitive to endogeneity concerns stemming from joint EHR decision making among physicians and hospitals.

The results suggest that EHR implementation may have intensified access to care issues for patients, and that there was a surge of experienced physicians who left the labor force and were no longer available to influence early career physicians. Physicians are also likely to change their behavior due to adverse work environments. Given the unintended consequences found here, policymakers should carefully consider how workers may respond to policies that affect their work environment.

@@

\subsection{Background on EHRs}
EHRs have been an important feature of health care since the 1980s. Early in the technology's existence, health care professionals perceived the technology as a complement to paper records, primarily deployed by large academic medical centers to improve efficiency in billing and/or scheduling. Physicians did not interact directly with these early-generation EHRs, and thus were not drastically affected by their implementation. As innovation made computers more portable, the usability of EHRs increased, creating what is known as the ``physician workspace": a computer station for a physician to interface directly with an EHR to record patient updates. Despite usability, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry. Automation in data entry was non-existent, making it extremely time consuming for the user.

The HITECH Act was passed in 2009, designating \$27 billion in government subsidies to entities who used EHRs according to certain guidelines. The guidelines included having at least 80\% of patients in the system, regularly recording answers to specific questions, and ensuring privacy. The U.S. allocated subsidies according to stages: stage 1 (2011) focused on data collection, stage 2 (2012) extended to using the EHR for care support, stage 3 (2014) extended to data sharing between practices. This program was successful in  spurring EHR adoption, shown in a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}). Figure \ref{fig:meanuse} shows a geographical comparison of U.S. hospitals who have received stage 1 of the meaningful use subsidy in 2011 vs. 2013, revealing the nationwide and rapid expansion of technology.

@@

On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively which are the most time consuming portions of EHR use (\cite{arndt2017tethered}). In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR, an example of which is shown in Figure \ref{fig:EPIC}. Despite a large amount of time spent with them, physicians still continue to report frustration over EHR use. The most time consuming functions are also reported as the most frustrating (\cite{dymek2021building}). A physician still likely spends almost 2 hours a day managing their inbox: deleting duplicate messages, sifting through messages meant for other members of a care team, and searching relevant information (\cite{dymek2021building}). Another common issue physicians face is the lack of usability of EHR systems, that functions either take minutes to locate or load. Hospitals often provide training upon adoption, but rapid system upgrades and changes make familiarity with the system difficult.

@@

This section serves to outline the potential responses of hospitalists to EHRs based on their incentives. The physician behaviors that I consider in this analysis are closely related and should not be taken as independent. For example, a physician's choice of work setting depends on their decision to prolong retirement, and billing activity depends on work setting. Thus, I discuss the behaviors in stages of decision making.

@@

Workers in developed countries tend to plan for formal retirement well in advance. Generally, an exogenous shock to a worker's environment is not expected to change retirement age due to the amount of planning necessary to formally leave the labor force. Classic retirement models structure the decision of retirement by maximizing expected lifetime utility based on future earnings and retirement benefits, or choosing to prolong retirement only if the expected earnings gain from doing so is positive (\cite{gustman1986disaggregated}, \cite{stock1990pension}). However, physicians typically make this decision differently than workers in other industries. Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}). When the time to retire comes, many physicians hesitate to abandon patients they have seen over the course of their career. The decision to delay retirement is not financial, but altruistic. Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.

It is important to note that age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently. First, I consider physicians already of retirement age ($>=60$) when a new technology is implemented. While payment schemes, wealth, and retirement benefits certainly impact retirement since, on average, those in this age group have already prepared financially to retire. Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility depends on job satisfaction and personal life factors. Exogenous EHR implementation affects the decision to retire through job satisfaction. If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work. If EHRs make daily tasks easier, there is a positive relationship. Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction. Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}).

Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role. There are opportunities for physicians of any age to switch careers towards administrative roles, teaching, consulting, or hospital management. While it is unlikely that early career physicians are formally retiring, physicians of any age could be career switching. This decision is impacted similarly to the above discussion regarding formal retirement, where a physician who is indifferent between clinical work and career switching can be affected one way or the other based on how an EHR impacts job satisfaction. As I discuss in Section \ref{sec:data}, I only observe whether a physician stops seeing Medicare patients, not whether they retire or career switch. Thus, the term retirement for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards. A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.

@@

For most physicians, EHRs will not impose a large enough cost (or benefit) to change the duration of working in clinical settings. However, there are other ways physicians might change their behavior under drastic changes to their current working environment. For those who remain practicing, changing the physical place of work may allow the physician more control over which technologies they use. Some hospitalists work under contract full time for one medical practice, where they have a consistent shift schedule, while others contract themselves out to multiple hospitals, offices, or both. While I do not have sufficient data to observe the type of hospitalist each physician is, I investigate the incentives faced by each type.

First, I consider those who are in a relatively binding contract with the hospital. For hospitalists under full time contract with a medical practice, the physician is more tethered to the hospital. This could impact the response to EHR implementation in that hospital in two ways: (1) through potentially increased use of the EHR in its initial phases, and (2) limited access to other work settings. A typical contract requires 90 to 120 days termination notice and some contracts include do not compete clauses (\cite{yasgur_by_-_yasgur_2016}), making a shift in work setting costly or excluding it from the choice set. Since these hospitalists are established in their hospital, they work a significant amount of time there. When an EHR is implemented, they may become familiar with it more quickly and be less likely to desire to switch work settings. Thus, I classify these physicians as ``sticky", since they are not able to change behavior in the short run.

@@

There are still physicians who are not induced to change their behavior on the basis of retirement or work setting due to EHR implementation. A lack of response could be due to 4 factors: the hospitalist is bound to their current work environment, EHRs make no impact on the work environment, EHRs are costly, but not costly enough to shift, or EHRs improve job satisfaction in the current work environment. For physicians who make no physical changes to work environment after EHR exposure, it is natural to assess whether EHRs affect productivity, as a main purpose of the HITECH Act was to improve efficiency of care, specifically to decrease time spent on administrative burden.

Whether EHRs affect productivity in terms of patients seen is theoretically ambiguous. If they serve their purpose to decrease administrative burden, the physician would be freed up to see more patients. In this case, productivity would increase. However, if utilizing an EHR actually imposes an additional administrative burden as suggested in physician interviews, the number of patients could either decrease or stay the same, indicating a decrease in productivity. Further, there is a unique relationship between EHR use and efficiency in terms of billing activity done by the physician. Billing activity refers to the number of claims filed per patient seen. EHRs may improve efficiency if they remove the need for repeated or excess testing. However, if having an EHR makes it easy for a physician to claim for additional items with the click of a button, the adverse incentive may lead physicians to claim for more items per patient.

@@

I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time, labormarket outcomes, and other relevant physician, hospital, and practice characteristics. I describe the different data sets used to construct the panel below, and I include a detailed outline of data and variable construction in Appendix \ref{app:data}.

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings. I only include physicians who reported their primary specialty as hospitalist (physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting), internal medicine, general practice or family practice in at least one year. Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 70\% of patients in a hospital setting in all years, and I drop physicians with insufficient claim counts to show indications of behavior. I use this data to construct the dependent variables for my analysis. Generally, these include the decision to retire (leave clinical work), switch to a different work setting, number of patients seen, and overall billing activity. I describe these outcome variables and their construction in detail in Section \ref{sec:outcome}.

CMS Shared Patient data records annual information on frequency of billing associated with the same patient across providers within a given time period. These data are available from 2009 to 2015. For example, if a primary care physician refers a patient to a specialist, then those two physicians have that particular shared patient in common. The number of shared patients are collected in 30, 90, or 180 day intervals. I focus on the 30 day interval, and I employ this data to detect physicians who work closely with hospitals. I limit the entities by tax-code to only include shared patients for physician-hospital pairs, where the physicians in these pairs match the physicians from MD-PPAS. The types of physicians are already limited to primary care, but there still may be office-based primary care physicians in the shared patient data. Therefore, I limit the sample of pairs based on a threshold of same day patients which depends on the number of years the physician appears in the data. I only include physician-hospital pairs who have non-missing data with at least 30 shared patients per year.

Using hospital National Provider Identifier (NPI), I link the physician-hospital pairs from the Shared Patient data to the American Hospital Association (AHA) survey, which contains information on hospital-level EHR use and other characteristics. I record the first year a physician is exposed to an EHR based on implementation in the hospitals they share patients with. A hospital can answer the question of whether they use an EHR with either ``Partially", ``Fully", or ``Not at all". Since partial implementation of an EHR does not guarantee sufficient use, I define a hospital as implementing an EHR the first year they answer ``Fully". I then aggregate this data to the physician level. Since the Shared Patient data does not extend to 2017 as the MD-PPAS data does, in the analysis I drop physicians who are not treated by 2015, as I do not observe whether they become treated in 2016 or 2017. That is, the data contains no units which are ``never-treated".

Finally, I include a physician's graduation year from Physician Compare. I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship. The details of such thresholds are explained in Appendix \ref{app:data}. To ensure that my results  are not sensitive to various levels of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations. The specification charts with these results are shown in Appendix \ref{sec:chart}, and I discuss each in conjunction with the main results.

@@

The first dependent variable I consider is the decision for a physician to stop seeing Medicare patients, which I refer to as retirement. This variable is constructed as follows: 
$$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$. Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future. \footnote{Alternatively, I could also define retirement using future number of patients. I use claim count for a more conservative measure of retirement, but using patient count yields identical results. In Appendix \ref{sec:changes}, I limit the sample years to 2009-2015 to assess the robustness of the main retirement definition.}
